---
title: "chapter3.Rmd"
author: "Leenapaivikki"
date: "17112018"
output: html_document
---

```{r setup, echo=FALSE, message=FALSE}
# Clear memory.
rm(list = ls())

# Packages required by this script.
library(dplyr)
library(GGally)
library(ggplot2)
library(boot)

```
#Analysis of the alcohol comsumption of Portugese students

#Introduction

The data of this analysis is from the Portugese survey on student performance in secondary education (high school). The database was built from two sources: school reports and questionnaires, used to complement the previous information. The data was collected during the years 2005-2006. The performance was scrutinized in two distinct subjects: Mathematics and Portuguese language.

The datasets were modelled by P. Cortez and A. Silva, described in the paper "Using Data Mining to Predict Secondary School Student Performance" (2008) http://www3.dsi.uminho.pt/pcortez/student.pdf

## Description of the data

#Data is read in
```{r data_input3}
alc <- as.data.frame(read.table('data/alc.csv',  sep="\t", header=TRUE))
```

